Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula

### 2024-02-12 output/paper/manuscript.qmd
Trying to fix Table 1, now with by cohort only in SM. Modifying names EDCs in forest plots.

format:
  docx:
    # reference-doc: reference_document.docx
    page-width: 8.3
    citeproc: true
    mermaid:
      theme: neutral
  # pdf:
  #   pdf-engine: lualatex
  #   number-sections: true
  #   toc: false
  #   keep-tex: false
  #   include-in-header:
  #     text: |
  #       \usepackage{lineno}
  #       \usepackage{rotating}
  #       \usepackage{float}
  #       \usepackage{typearea}
  #   include-before-body:
  #     text: |
  #       \linenumbers

@@

filters:
  - acronyms
  # - authors-block
  # - ../../../../all_phd/_extensions/wordcount/citeproc.lua
  - ../../../../all_phd/_extensions/wordcount/wordcount.lua

@@

  val <- df[df$variable == exposure & df$outcome == outcome, ]$val
  val <- stringr::str_replace(
    string = val,
    pattern = "\\*",
    replacement = ""
  )
  glue::glue(
    "\{{< acr mc >}} and $95\\%$ \{{< acr ci >}}: {val}"
  )
}
```

@@

Epidemiological evidence suggests that exposure to {{< acr edcs >}} may be linked to perturbation of the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention measures in children. We used data from the {{< acr helix >}} cohort including 1,297 children aged 6-11. We used the parametric g-formula and {{< acr mcs >}} to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of efficiency from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed positive {{< acr mc >}}s for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with positive {{< acr mc >}}s of cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found possible effect modification by sex for cortisone and corticosterone production. Our results suggest that multiple {{< acr edcs >}} might interfere with inattentiveness outcomes in children and with the homeostasis of the {{< acr hpa >}} axis. Further research is necessary to determine whether this can have a clinically significant impact on childhood neurodevelopment.

@@

{{< acr opps >}} are one group of {{< acr edcs >}} that may have a deleterious effect on neurodevelopment, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size and few {{< acr opp >}} metabolites measured [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites during childhood and early adolescence has been associated to several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to {{< acr bpa >}} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

@@

Taken together, these results suggest that the negative influence of exposure to certain {{< acr edcs >}} on neurodevelopmental outcomes, might be mediated, at least partially, by disruption of the {{< acr hpa >}} axis' homeostasis.
In the present study, we thus estimated associations between 1) non-persistent {{< acr edcs >}} and attention, 2) non-persistent {{< acr edcs >}} and glucocorticosteroids, and 3) glucocorticosteroids and attention, using the parametric g-formula and {{< mc >}}s, in children of a large cohort in Europe.

@@

We estimated {{< acr mc first_use=true >}}s with the parametric g-formula, a method of standardization. The parametric g-formula involves the following steps: 1) fit a outcome model including both covariates and balancing weights; 2) create two new datasets identical to the original one but with the exposure shifted according to a user-specified intervention set by a deterministic function of the observed exposure levels; 3) use the outcome model to compute adjusted predictions in the two counterfactual datasets; 4) compute the difference between the means of the adjusted predictions in the counterfactual datasets. The causal parameter of interest $\phi$ was thus specified as the difference in the expected counterfactual outcomes under the shifted exposure levels $\left( \phi^\Delta = \mathbb{E >}} \left[ Y^{d_1 >}} \right] - \mathbb{E >}} \left[ Y^{d_2 >}} \right] \right)$. In order for this parameter to be identified, the usual causal identifiability conditions (no unmeasured confounding, positivity, and consistency) are required. Since these conditions are likely not satisfied, we focused on the estimation of a statistical estimand that is as close as possible to the causal parameter of interest.

@@

To estimate the {{< acr mc >}}s, we used the `avg_comparisons` function from the `marginaleffects` R package [@Arel-Bundock:2023]. The two counterfactual datasets were obtained by setting the exposures levels to 90th percentile ($d_1$) and the 10th percentile ($d_2$), for each cohort separately. The {{< acr mc >}}s were computed using the estimated balancing weights above. Robust standard errors were computed with the `sandwich` R package, using cohort as variable indicating clustering of observations [@Zeileis:2004; @ZeileisKollGraham:2020].
For each outcome, we report the results as differences between {{< acr mc >}}s.

We further estimated the {{< acr adrf >}} using the `avg_predictions` function from the `marginaleffects` R package, examining 50 exposure values from the 10th to the 90th percentiles of the exposure. As done for the {{< acr mc >}}s, we included the estimated balancing weights and used cohort as a clustering variable when computing robust standard errors.

@@

We tested for possible effect-modification by sex. To do so, balancing weights were estimated separately for each level of the sex variable, and an interaction term between the exposure and sex was included in the outcome model. Similarly, the {{< acr mc >}}s were aggregated separately for each level of sex. We further tested for significance of the difference between the {{< acr mc >}}s of males and females.

@@

Of the 1,301 children of the {{< acr helix >}} subcohort, 1,297 had measurements of the non-persistent {{< acr edcs >}}. Measurements of the glucocorticosteroids were available for 1,004 children, of which 980 were matched to the {{< acr helix >}} subcohort. Measurements of both non-persistent {{< acr edcs >}} and glucocorticosteroids were available for 976 children of the subcohort. A flowchart describing the sample size for each research question is presented in @suppfig-flow-pop. The sample consisted of 55% girls. The median {{< acr hitrtse >}} was 300 ms ({{< acr iqr >}}, 231-368), with lower median values for EDEN, MOBA, and INMA, corresponding to the cohorts with older children. At the time of visit, the median age of the children was 8.06 years. The children were mostly Caucasian (90%), and the largest minority were of Pakistani origin (6.2%). @tbl-pop-desc provides descriptive statistics for the outcome and covariates, by cohort and overall.

@@

@Fig-marginal-1 presents the forest plot for the {{< acr mc >}}s of the non-persistent {{< acr edcs >}} on {{< acr hitrtse >}}. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the values of {{< acr hitrtse >}}. Most of the {{< acr ci first_use=true >}}s included the null effect, though. Statistically significant effects were observed for the phenol MEPA (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`). The {{< acr opp >}} DETP was strongly negatively associated with {{< acr hitrtse >}} (`r format_res(meffects1, "detp", "hitrtse")`).

@Fig-marginal-2 presents the forest plot for the {{< acr mc >}}s of the non-persistent {{< acr edcs >}} on cortisone production, cortisol production, and corticosterone production. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the total production of these metabolites. Exceptions were BUPA, which was associated with negative {{< acr mc >}}s for all three outcomes, and MiBP, which was associated with a negative {{< acr mc >}} for total cortisone production only.
The majority of the effects for the phenols and phthalate metabolites included the null. The phenol BPA showed the largest {{< acr mc >}}s across all three outcomes (cortisone production, `r format_res(meffects2, "bpa", "cortisone production")`; cortisol production, `r format_res(meffects1, "bpa", "cortisol production")`; corticosterone production, `r format_res(meffects1, "bpa", "corticosterone production")`).

@Fig-marginal-3 presents the forest plot for the {{< acr mc >}}s of the glucocorticosteroids on {{< acr hitrtse >}}. All {{< acr mc >}}s included the null, with no clear indication of directionality of the effect.

@@

@Tbl-hypothesis-1 presents the results of a hypothesis test for the difference between estimates of the {{< acr mc >}}s for females and males, for each exposure. Significant differences were present for the phenol OXBE (`r format_res(heffects1, "oxbe", "hitrtse")`) and the phthalate metabolites MEP (`r format_res(heffects1, "mep", "hitrtse")`) and MbZP (`r format_res(heffects1, "mbzp", "hitrtse")`). The forest plot of the individual {{< acr mc >}}s is presented in @suppfig-marginal-1sa.

@Tbl-hypothesis-2 presents the results of a hypothesis test for the difference between estimates of the {{< acr mc >}}s for females and males, for each exposure and outcome. Significant differences were present across all three classes of {{< acr edcs >}} and for all outcomes. The largest differences were attributable to the {{< acr opps >}} DMTP (cortisol production, `r format_res(heffects2, "dmtp", "cortisol production")`) and DETP (corticosterone production, (`r format_res(heffects2, "detp", "corticosterone production")`); cortisone production, (`r format_res(heffects2, "detp", "cortisone production")`)). The forest plots of the individual {{< acr mc >}}s are presented in @suppfig-marginal-2sa.

@Tbl-hypothesis-3 presents the results of a hypothesis test for the difference between estimates of the {{< acr mc >}}s for females and males, for each exposure. Significant differences were present for cortisone production (`r format_res(heffects3, "cortisone prod.", "hitrtse")`) and corticosterone production (`r format_res(heffects3, "corticost. prod.", "hitrtse")`). Furthermore, for all exposures, the {{< acr mc >}}s had opposite sign (positive for males and negative for females). The forest plot of the individual {{< acr mc >}}s is presented in @suppfig-marginal-3sa.

@@

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent {{< acr edcs >}} had negative effects on {{< acr hitrtse >}} and total production of cortisol, cortisone, and corticosterone, although the majority of the {{< acr ci >}}s included the null effect. Increased production of these glucocorticosteroids did not seem to affect {{< acr hitrtse >}}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to {{< acr edcs >}} in relation to {{< acr hitrtse >}}. The literature on {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to {{< acr adhd >}} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of {{< acr edcs >}} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with {{< acr adhd >}} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several {{< acr edcs >}} on neurodevelopment in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects are stronger in males. A major limitation of these studies is the reliance on spot urine samples, that might not be representative of long-term exposures.

@@

Strengths of the present study include the use of pooled urine samples for chemical assessment, since it is known that these specific {{< acr edcs >}} have very short half-lives [@CasasBasaganaSakhi:2018].
We decided to model both the *treatment* mechanisms, for the estimation of balancing weights, and the outcomes, with traditional covariates adjustment, to try to obtain *doubly robust* effect estimates. Finally, we decided not to interpret our results by focusing on the estimated coefficients of the regression models, but by making use of the g-computation procedure and estimate {{< acr mc >}}s.

>>>>

format:
  docx:
    # reference-doc: reference_document.docx
    page-width: 8.3
    citeproc: true
    mermaid:
      theme: neutral
  pdf:
    documentclass: article
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    pdf-engine: lualatex
    number-sections: true
    toc: false
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{lineno}
        \usepackage{rotating}
        \usepackage{float}
        \usepackage{typearea}
    include-before-body:
      text: |
        \linenumbers

@@

filters:
  - acronyms
  - ../../../../all_phd/_extensions/wordcount/wordcount.lua

@@

  val <- df[df$variable == exposure & df$outcome == outcome, ]$val
  val <- stringr::str_replace(
    string = val,
    pattern = "\\*",
    replacement = ""
  )
  glue::glue(
    "\\acr{{mc}} and $95\\%$ \\acr{{ci}}: {val}"
  )
}
```

@@

Epidemiological evidence suggests that exposure to {{< acr edcs >}} may be linked to perturbation of the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention measures in children. We used data from the {{< acr helix >}} cohort including 1,297 children aged 6-11. We used the parametric g-formula and {{< acr mcs >}} to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of efficiency from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed positive {{< acr mcs >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with positive {{< acr mcs >}} of cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found possible effect modification by sex for cortisone and corticosterone production. Our results suggest that multiple {{< acr edcs >}} might interfere with inattentiveness outcomes in children and with the homeostasis of the {{< acr hpa >}} axis. Further research is necessary to determine whether this can have a clinically significant impact on childhood neurodevelopment.

@@

One group of {{< acr edcs >}} that may have a deleterious effect on neurodevelopment is the {{< acr opps >}}, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size and few {{< acr opp >}} metabolites measured [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites during childhood and early adolescence has been associated to several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to {{< acr bpa >}} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

@@

Taken together, these results suggest that the negative influence of exposure to certain {{< acr edcs >}} on neurodevelopmental outcomes, might be mediated, at least partially, by disruption of the {{< acr hpa >}} axis' homeostasis.
In the present study, we thus estimated associations between 1) non-persistent {{< acr edcs >}} and attention, 2) non-persistent {{< acr edcs >}} and glucocorticosteroids, and 3) glucocorticosteroids and attention, using the parametric g-formula and {{< acr mcs first_use=true >}}, in children of a large cohort in Europe.

@@

We estimated {{< acr mcs >}} with the parametric g-formula, a method of standardization. The parametric g-formula involves the following steps: 1) fit a outcome model including both covariates and balancing weights; 2) create two new datasets identical to the original one but with the exposure shifted according to a user-specified intervention set by a deterministic function of the observed exposure levels; 3) use the outcome model to compute adjusted predictions in the two counterfactual datasets; 4) compute the difference between the means of the adjusted predictions in the counterfactual datasets. The causal parameter of interest $\phi$ was thus specified as the difference in the expected counterfactual outcomes under the shifted exposure levels $\left( \phi^\Delta = \mathbb{E} \left[ Y^{d_1} \right] - \mathbb{E} \left[ Y^{d_2} \right] \right)$. In order for this parameter to be identified, the usual causal identifiability conditions (no unmeasured confounding, positivity, and consistency) are required. Since these conditions are likely not satisfied, we focused on the estimation of a statistical estimand that is as close as possible to the causal parameter of interest.

@@

To estimate the {{< acr mcs >}}, we used the `avg_comparisons` function from the `marginaleffects` R package [@Arel-Bundock:2023]. The two counterfactual datasets were obtained by setting the exposures levels to 90th percentile ($d_1$) and the 10th percentile ($d_2$), for each cohort separately. The {{< acr mcs >}} were computed using the estimated balancing weights above. Robust standard errors were computed with the `sandwich` R package, using cohort as variable indicating clustering of observations [@Zeileis:2004; @ZeileisKollGraham:2020].
For each outcome, we report the results as differences between {{< acr mcs >}}.

We further estimated the {{< acr adrf >}} using the `avg_predictions` function from the `marginaleffects` R package, examining 50 exposure values from the 10th to the 90th percentiles of the exposure. As done for the {{< acr mcs >}}, we included the estimated balancing weights and used cohort as a clustering variable when computing robust standard errors.

@@

We tested for possible effect-modification by sex. To do so, balancing weights were estimated separately for each level of the sex variable, and an interaction term between the exposure and sex was included in the outcome model. Similarly, the {{< acr mcs >}} were aggregated separately for each level of sex. We further tested for significance of the difference between the {{< acr mcs >}} of males and females.

@@

@tbl-pop-desc and @supptbl-pop-desc provide descriptive statistics for the outcome and covariates for the {{< acr helix >}} subcohort and for each cohort, respectively.
Of the 1,301 children of the {{< acr helix >}} subcohort, 1,297 had measurements of the non-persistent {{< acr edcs >}}. Measurements of the glucocorticosteroids were available for 1,004 children, of which 980 were matched to the {{< acr helix >}} subcohort. Measurements of both non-persistent {{< acr edcs >}} and glucocorticosteroids were available for 976 children of the subcohort. A flowchart describing the sample size for each research question is presented in @suppfig-flow-pop. The sample consisted of 55% girls. The median {{< acr hitrtse >}} was 300 ms ({{< acr iqr >}}, 231-368), with lower median values for EDEN, MOBA, and INMA, corresponding to the cohorts with older children. At the time of visit, the median age of the children was 8.06 years. The children were mostly Caucasian (90%), and the largest minority were of Pakistani origin (6.2%).

@@

@Fig-marginal-1 presents the forest plot for the {{< acr mcs >}} of the non-persistent {{< acr edcs >}} on {{< acr hitrtse >}}. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the values of {{< acr hitrtse >}}. Most of the {{< acr cis first_use=true >}} included the null effect, though. Statistically significant effects were observed for the phenol MEPA (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`). The {{< acr opp >}} DETP was strongly negatively associated with {{< acr hitrtse >}} (`r format_res(meffects1, "detp", "hitrtse")`).

@Fig-marginal-2 presents the forest plot for the {{< acr mcs >}} of the non-persistent {{< acr edcs >}} on cortisone production, cortisol production, and corticosterone production. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the total production of these metabolites. Exceptions were BUPA, which was associated with negative {{< acr mcs >}} for all three outcomes, and MiBP, which was associated with a negative {{< acr mc >}} for total cortisone production only.
The majority of the effects for the phenols and phthalate metabolites included the null. The phenol BPA showed the largest {{< acr mcs >}} across all three outcomes (cortisone production, `r format_res(meffects2, "bpa", "cortisone production")`; cortisol production, `r format_res(meffects2, "bpa", "cortisol production")`; corticosterone production, `r format_res(meffects2, "bpa", "corticosterone production")`).

@Fig-marginal-3 presents the forest plot for the {{< acr mcs >}} of the glucocorticosteroids on {{< acr hitrtse >}}. All {{< acr mcs >}} included the null, with no clear indication of directionality of the effect.

@@

@Tbl-hypothesis-1and2 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for each exposure. Significant differences were present for the phenol OXBE (`r format_res(heffects1, "oxbe", "hitrtse")`) and the phthalate metabolites MEP (`r format_res(heffects1, "mep", "hitrtse")`) and MbZP (`r format_res(heffects1, "mbzp", "hitrtse")`). The forest plot of the individual {{< acr mcs >}} is presented in @suppfig-marginal-1sa.

@Tbl-hypothesis-1and2 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for each exposure and outcome. Significant differences were present across all three classes of {{< acr edcs >}} and for all outcomes. The largest differences were attributable to the {{< acr opps >}} DMTP (cortisol production, `r format_res(heffects2, "dmtp", "cortisol production")`) and DETP (corticosterone production, (`r format_res(heffects2, "detp", "corticosterone production")`); cortisone production, (`r format_res(heffects2, "detp", "cortisone production")`)). The forest plots of the individual {{< acr mcs >}} are presented in @suppfig-marginal-2sa.

@Tbl-hypothesis-3 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for each exposure. Significant differences were present for cortisone production (`r format_res(heffects3, "cortisone prod.", "hitrtse")`) and corticosterone production (`r format_res(heffects3, "corticost. prod.", "hitrtse")`). Furthermore, for all exposures, the {{< acr mcs >}} had opposite sign (positive for males and negative for females). The forest plot of the individual {{< acr mcs >}} is presented in @suppfig-marginal-3sa.

@@

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent {{< acr edcs >}} had negative effects on {{< acr hitrtse >}} and total production of cortisol, cortisone, and corticosterone, although the majority of the {{< acr cis >}} included the null effect. Increased production of these glucocorticosteroids did not seem to affect {{< acr hitrtse >}}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to {{< acr edcs >}} in relation to {{< acr hitrtse >}}. The literature on {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to {{< acr adhd >}} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of {{< acr edcs >}} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with {{< acr adhd >}} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several {{< acr edcs >}} on neurodevelopment in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects are stronger in males. A major limitation of these studies is the reliance on spot urine samples, that might not be representative of long-term exposures.

@@

Strengths of the present study include the use of pooled urine samples for chemical assessment, since it is known that these specific {{< acr edcs >}} have very short half-lives [@CasasBasaganaSakhi:2018].
We decided to model both the *treatment* mechanisms, for the estimation of balancing weights, and the outcomes, with traditional covariates adjustment, to try to obtain *doubly robust* effect estimates. Finally, we decided not to interpret our results by focusing on the estimated coefficients of the regression models, but by making use of the g-computation procedure and estimate {{< acr mcs >}}.
